---
title: "Using the Mosaic Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Mosaic Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Basic steps

_mosaic_ has no strict dependencies, so you can get started right away.

First, load up _mosaic_, create some coordinate data falling in a ring-shaped pattern using _mosaic_'s _ptpattern_ function, and stash them in x and y arrays:

```{r}
library(mosaic)
p <- ptpattern(50,shape='ring')

x <- p$x
y <- p$y
```

Make a quick plot to get an idea of the density of sampling, and check the ranges. The data are centered on the origin, with a default maximum range of -0.5 to 0.5. This can be changed by setting the diameter parameter, which defaults to 1.

```{r fig.height=5, fig.width=5}
plot(x,y,pch=19)
range(x)
range(y)
```

We'll go back to the x and y data momentarily, but note that _ptpattern_ will produce other shapes: circle, square, and rectangle. If 'rectangle' is chosen, the default aspect ratio is 2, meaning that the x-axis data range from -1 to 1. So, to create a rectangle 1 unit wide and 10 units high, use both the _diameter_ and the _asp_ parameters:

```{r fig.height=5, fig.width=5}
p <- ptpattern(points=20,shape='rectangle',diameter=10,asp=0.1)
plot(p$x,p$y,pch=19)
```

Note the x and y axis limits.

The main purpose of _mosaic_ is to estimate areas of shapes like the ring shown above. _tgraph_ bases its estimates on minimum spanning trees (MSTs) and _mgraph_ bases its on MSTs combined with mosaic graphs.

Going back to the data, the areas are easy to extract:

```{r}
t <- tgraph(x,y)
m <- mgraph(x,y)
t$area
m$area
```

## Plotting mosaics and trees

Both functions use _plot.mosaic_ to present graphs. First plot a simple mosaic:

```{r fig.height=4, fig.width=5}
m <- mgraph(x,y)
plot(m)
```

See how _plot.mosaic_ fills up the entire space and omits superfluous axes and axis labels.

The overall MST doesn't need to match up exactly with the mosaic, but can come close. As you might expect, _plot.mosaic_ respects add=T and lwd:

```{r fig.height=4, fig.width=5}
plot(m)
plot(t,add=T,lwd=2)
```

So if you really wanted those axes and labels, you could plot the data using some other plot method and then add the mosaic or MST using add=T. Note that axis() and mtext() wouldn't work because the figure's margins are set to zero automatically.

_plot.mosaic_ has numerous options. You can specify colors by entering either a _col_ value like "red" or "blue" or an _hsv_ specification. In the latter case, colors are set by entering h, s, v, and alpha parameters. _plot.mosaic_ defaults to a light blue color (h = 0.5), zero saturation (s = 0), medium darkness (v = 0.5), and halfway transparency (alpha = 0.5), so to get any non-black color at all you need to ramp up the saturation. Lightening things up with v=1 also helps. Also, _plot.mosaic_ needs to be told to color the lines and points by toggling _pts.black_ and _lines.black_. The hue of red is about 0.05. We'll omit the lwd=2 part because colors will now differentiate the graphs.

```{r fig.height=4, fig.width=5}
plot(m,h=0.05,s=1,v=1,pts.black=F,lines.black=F)
```

Now add the MST in green (h = 0.4), stripping out the points in the second call to _plot.mosaic_ because they're shown already. This is done by setting cex to zero.

```{r fig.height=4, fig.width=5}
plot(m,h=0.05,s=1,v=1,pts.black=F,lines.black=F)
plot(t,add=T,h=0.4,s=1,v=1,lines.black=F,cex=0)
```

_mgraph_ bases its estimates on the square of summed edge lengths. To visualize the "area" contribution of each edge, use the _squares_ option. Leave the points and lines black so you can tell them apart from the squares.

```{r fig.height=4, fig.width=5}
plot(m,h=0.05,s=1,v=1,squares=T)
```

_plot.mosaic_ also allows you to visualize the data by filling in the hull of the mosaic. This isn't a convex hull, because it strictly follows the outermost edges in the graph. _plot.mosaic_ automatically uses a utility function called _mhull_ to do this; _mhull_ is fast, so the process is seamless.

```{r fig.height=4, fig.width=5}
plot(m,h=0.05,s=1,v=1,fill=T,lines=F)
```

_plot.mosaic_ can also just draw an outline to show the hull, but it will only be easily visible if you suppress the edge lines using lines=F. Let's try purple (h = 0.8) this time.

```{r fig.height=4, fig.width=5}
plot(m,h=0.8,s=1,v=1,outline=T,lines=F)
```

Another possibility is to add an MST. We'll use a square shape instead of a ring shape because MSTs are boring when the data fall in a ring. First, here's the basic mosaic shape. Note how _mgraph_ and _tgraph_ can be embedded within the _plot_ call.


```{r fig.height=4, fig.width=5}
p <- ptpattern(50)
plot(mgraph(p$x,p$y),h=0.8,s=1,v=1,lines.black=F)
```

Second, the filled mosaic with the MST on top (in yellow: h = 0.15). Note that the MST doesn't have to follow the mosaic, so it can even jump outside of the filled hull. Drop the points, and don't forget lines.black=F and add=T:

```{r fig.height=4, fig.width=5}
plot(mgraph(p$x,p$y),cex=0,h=0.8,s=1,v=1,fill=T,lines=F)
plot(tgraph(p$x,p$y),cex=0,h=0.15,s=1,v=1,lwd=2,lines.black=F,add=T)
```

_mgraph_ will break up the data into separate networks depending on how the _neighbors_ and _mutual_ options are set. By default, if all of the points in one region have nearest neighbors in that region alone, and likewise the points in other regions, then no edges will be drawn between the regions. To put this more simply, if the data form tight clusters then the clusters will be kept separate.

This is nothing to worry about because the routines for estimating area and for plotting the data handle separate clusters automatically. Here's an example showing two disconnected subgraphs.

```{r fig.height=4, fig.width=5}
p1 <- ptpattern(50)
p2 <- ptpattern(50)
x <- c(p1$x,p2$x + 2)
y <- c(p1$y,p2$y + 2)
plot(mgraph(x,y),cex=0.5,h=0.15,s=1,v=1,fill=T,lines=F)
```

To summarize, _mgraph_ provides x and y coordinates in addition to area estimates. It also can tell you the length of the mosaic, the distances between points, and the structure of the graph in terms of a binary, triangular matrix, where 1 = connected and 0 = disconnected:

```{r}
m <- mgraph(runif(10),runif(10))
m$length
m$distances
m$graph
```

Some users may wish to explot the filled mosaics for use in packages like _sf_, _sp_, and _raster_. The function _mhull_ provides a polygon in simple features format, which _sf_ understands natively:

```{r}
p <- mhull(m)$polygon
p
```

This means you can use handy _sf_ functions like _st_area_ without any hassle. Actually, I don't recommend using _st_area_ because the whole point of this package is to provide accurate area estimates, and it only creates polygons for illustrative purposes. But that's still an option.

Note that if _mgraph_ breaks up the data into more than one subgraph, meaning disconnected networks, _mhull_ will export a batch of polygons bound together as a list. Again, _sf_ will understand immediately what this means, so you don't have to do anything in particular.

It's also easy to get the data into a format that _sp_ will understand using the _sf_ functions _st_sfc_ and _as_Spatial_, as in:

p <- as_Spatial(st_sfc(p))

And you can export the data to a shapefile using the _raster_ function _shapefile_:

shapefile(as_Spatial(st_sfc(p)),'mosaic.shp')

## Plotting GBIF data

This section shows how to draw data for an individual species from the Global Biodiversity Information Facility, draw a map, and plot a mosaic on the map. Fortunately, the _rgbif_ and _maps_ libraries make this easy. To get rid of any obviously bogus coordinates, we'll use the _CoordinateCleaner_ package.

```{r}
library(CoordinateCleaner)
library(maps)
library(rgbif)
```

The geographic range of the humble yellow-spotted rock hyrax (_Heterohyrax brucei_) is a representative example of GBIF data. The _rgbif_ function _occ_search_ has lots of options, but most aren't important for a simple example. However, it's useful to throw out the records lacking coordinates, and museum collection data are more reliable (basisOfRecord='PRESERVED_SPECIMEN'). _occ_search_ returns a big complicated object, of which only the third list item is needed. We'll convert that to a data frame. The actual coordinates can be backed out of it easily.

```{r}
species <- 'Heterohyrax brucei'
o <- occ_search(scientificName=species,hasCoordinate=T,basisOfRecord='PRESERVED_SPECIMEN')
d <- data.frame(o[3])
lat <- d$data.decimalLatitude
long <- d$data.decimalLongitude
```

_CoordinateCleaner_ has an option to remove zero values that's too aggressive, but the other ones are useful. It returns a matrix of true and false values (= passed or failed the test), so we'll mark any row with an F as a bad row. Finally, the lat and long values will be trimmed down by only taking the good ones.

```{r}
x <- data.frame(long,lat,species)
cc <- clean_coordinates(x,lon='long',lat='lat',species='species',
 tests=c("capitals","centroids","gbif","institutions"))

bad <- array(dim=length(lat),data=0)
for (j in 1:nrow(cc))
	if (sum(cc[j,] == F) > 0)	{
		bad[j] <- 1
	}
lat <- lat[bad == 0]
long <- long[bad == 0]
```

At this point, plotting the map becomes trivial. The Albers projection is a pretty good one. Most of the _map_ options are self-explanatory. It helps to set zero margins so the map takes up more room. lforce='s' is used to restrict the part of the map that is plotted; it's not always needed.


```{r fig.height=5.5, fig.width=5}
par(mar=c(0,0,0,0))
map(project='albers',parameters=c(0,0),xlim=c(-17,57),ylim=c(-30,30),fill=T,col='gray90',
 border=NA,lforce='s')
points(long,lat,cex=0.4,pch=21,col='black',bg='white')
```

Now it's easy to add a filled mosaic. For ease of reading, first get the graph object and then plot it. longlat = T tells _mgraph_ that distances between points should be great circle distances, not Euclidean distances. h = 0.65 sets the color to a purplish blue; alpha = 0.35 makes the filled mosaic mostly transparent.

```{r fig.height=5.5, fig.width=5}
par(mar=c(0,0,0,0))
map(project='albers',parameters=c(0,0),xlim=c(-17,57),ylim=c(-30,30),fill=T,col='gray90',
 border=NA,lforce='s')
m <- mgraph(long,lat,longlat=T)
plot(m,cex=0.4,h=0.65,s=1,v=1,alpha=0.35,fill=T,outline=T,lines=F,add=T)
```

Note that the mosaic includes several thin lines. These are bridges between mosaic pieces, and they are a totally normal thing to see when a data set is sparse.

Finally, another good way to visualize the data is to delete the fill and add squares that follow the edges.

```{r fig.height=5.5, fig.width=5}
par(mar=c(0,0,0,0))
map(project='albers',parameters=c(0,0),xlim=c(-17,57),ylim=c(-30,30),fill=T,col='gray90',
 border=NA,lforce='s')
plot(m,cex=0.4,h=0.65,s=1,v=1,alpha=0.35,lines=F,squares=T,add=T)
```

What about the area estimates, which are the whole point of the exercise? The mosaic and MST values are similar, but certainly not identical. The values are in square degrees.

```{r}
mgraph(long,lat,longlat=T)$area
tgraph(long,lat,longlat=T)$area
```

## Simulation analysis

Mosaic areas have three main advantages: they are accurate, insensitive to variation in sample size, and able to handle complex shapes. In this section, we'll contrast mosaic and MST areas with areas based on convex hulls (= minimum convex polygons) and on 95% kernel density estimates (KDEs).

We'll skip hypervolumes because the computations are very slow. But for the record, you can compute hypervolumes easily by loading the library _hypervolume_ and calling _get_volume(hypervolume_gaussian(pts))_. 

Note that the areas will be of circles (the default shape used by _ptpattern_).

To start with, we'll need the _sf_ library to calculate the hull areas and _MASS_ to get KDEs:

```{r}
library(MASS)
library(sf)
```

Differences among methods are pretty visible at sample sizes of around 10, and only about 1000 trials are needed to see fine-scale differences. Start by setting these parameters and arrays:

```{r}
points <- 10
trials <- 1000
hull_area <- array(dim=trials,data=NA)
kde_area <- array(dim=trials,data=NA)
mosaic_area <- array(dim=trials,data=NA)
mst_area <- array(dim=trials,data=NA)

```

The D array stores random shape diameters that fall over a broad, log-normally distributed range, with a mean of one linear unit (zero log units) and a standard deviation of 0.5. The true area values follow directly from the diameters, although they're also computed by the _ptpattern_ function. Note if _r_ is the radius and _D_ is the diameter, pi _r_^2 is the same as pi _D_^2/4.

```{r}
D <- exp(rnorm(trials,0,0.5))
true_area <- array(dim=trials,data=pi * D^2 / 4)
```

Getting the data in shape for the _sf_ area calculation is a little longwinded. The base function _chull_ is used to find the points that form the convex hull and put them in order; the first point is added to the end of the matrix so the hull will wrap around completely; _st_polygon_ and _st_sfc_ are used to prep the data; and _st_area_ is only used at the very end. Here's an example:

```{r}
p <- ptpattern(points=points,diameter=D[1])
pts <- cbind(p$x,p$y)

ch <- pts[chull(pts),]
ch <- rbind(ch,ch[1,])
p <- st_sfc(st_polygon(list(ch)))
st_area(p)
```

Meanwhile, the KDE calculations are a little complicated. First we need to get the mean and range of the data across each axis. The reason is that the _kde2d_ function defaults to only calculating densities over the narrowest possible range. Note that _kde2d_ spits out these values for a grid of cells.

```{r}
mx <- mean(pts[,1])
my <- mean(pts[,2])
rx <- 10 * diff(range(pts[,1]))
ry <- 10 * diff(range(pts[,2]))
```

In the following lines, the densities are obtained over a wide range using _kde2d_; the area of each grid cell used by _kde2d_ is calculated; the densities are sorted, standardized so they add up to 1, and cumulated; the cells making up the top 95% of the distribution are counted; and the count is scaled up with the area value ("cell").

```{r}
k <- kde2d(pts[,1],pts[,2],lims=c(mx - rx,mx + rx,my - ry,my + ry))
cell <- diff(k$x)[1] * diff(k$y)[1]
z <- sort(k$z,decreasing=T)
z <- z / sum(z)
z <- cumsum(z)
(sum(z < 0.95) + 1) * cell
```

By contrast, _mgraph_ and _tgraph_ do all the work for you:

```{r}
mgraph(pts)$area
tgraph(pts)$area
```

Finally, we're ready to loop through.

```{r}
for (i in 1:trials)	{
	p <- ptpattern(points=points,diameter=D[i])
	pts <- cbind(p$x,p$y)

	ch <- pts[chull(pts),]
	ch <- rbind(ch,ch[1,])
	p <- st_sfc(st_polygon(list(ch)))
	hull_area[i] <- st_area(p)

	mx <- mean(pts[,1])
	my <- mean(pts[,2])
	rx <- 10 * diff(range(pts[,1]))
	ry <- 10 * diff(range(pts[,2]))

	k <- kde2d(pts[,1],pts[,2],lims=c(mx - rx,mx + rx,my - ry,my + ry))
	cell <- diff(k$x)[1] * diff(k$y)[1]
	z <- sort(k$z,decreasing=T)
	z <- z / sum(z)
	z <- cumsum(z)
	kde_area[i] <- (sum(z < 0.95) + 1) * cell

	mosaic_area[i] <- mgraph(pts)$area
	mst_area[i] <- tgraph(pts)$area
}
```

The following two lines spit out the geometric mean of the ratio of estimated to true area. These simple statistics make the point pretty clearly.

```{r}
d <- data.frame(hull_area,kde_area,mst_area,mosaic_area) / true_area
exp(colMeans(log(d)))
```

There are lots of ways to illustrate the differences in accuracy and precision between the methods, such as a simple box plot:

```{r fig.height=4, fig.width=6}
boxplot(d)
abline(1,0)
```

Note that an accurate method would center the data on a value of 1 (meaning the ratio of the estimated to true area is 1). Mosaic and MST areas do this; the other two methods don't.

This plot is deceptive because the low hull area values are scrunched together. Log transforming the data makes more sense. Now 0 = perfectly accurate:

```{r fig.height=4, fig.width=6}
boxplot(log(d))
abline(0,0)
```

You can see that differences in precision (the size of the boxes) are small, but convex hulls are a bit noisier. Accuracy is the real issue.

A scatter plot makes the high accuracy of mosaic areas easier to grasp:

```{r fig.height=4, fig.width=5}
plot(true_area,mosaic_area,pch=19,cex=0.3,col='red',log='xy')
lines(c(0.01,100),c(0.01,100))
```

Staggering the data for each method by a factor of, say, 10 helps to compare them:

```{r fig.height=6, fig.width=5}
plot(true_area,mosaic_area,pch=19,cex=0.3,col='red',log='xy',ylim=c(0.01,10000),xlab='true area',ylab='area estimate')
points(true_area,10 * mst_area,pch=19,cex=0.3,col='blue')
points(true_area,100 * kde_area,pch=19,cex=0.3,col='green3')
points(true_area,1000 * hull_area,pch=19,cex=0.3,col='purple')
lines(c(0.01,100),c(0.01,100))
lines(c(0.01,100),10 * c(0.01,100))
lines(c(0.01,100),100 * c(0.01,100))
lines(c(0.01,100),1000 * c(0.01,100))
```

The lines are lines of unity, not regression lines, so not falling on them is a real problem. Specifically, the KDE values (green) are so high and the convex hull values (purple) are so low that they overlap heavily, despite the offsetting!


